---
title: "Explore_solutions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore_solutions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Solution explorer tool
# Evaluate tradeoffs across solution metrics
Here we can evaluate tradeoffs across metrics other than footprint size and frequency
of planning unit selection across all solutions. For example, we can understand the
amount of a solution within existing conservation lands, proportion of a solution that
occurs within fenced areas or native habitat.

There must be a file with metrics of interest associated with each planning unit
in order to run this tool. These values can be calculated using zonal statistics or intersections.
See the example dataset for setup: PU_metrics.gpkg

## Load libraries for running the explorer tool

```{r Prep, echo=TRUE, warning=FALSE, message=FALSE}
library(optimTFE)
library(optimTFE.explorer)
```

## Gather datasets needed.
This includes the pu_data spatial file with the planning unit id column and the 
solution summary metrics .csv file created from the 'all_sols_metrics' object in
the optimTFE vignette. Below also demonstrates how to create the files.

```{r Create planning unit file and solution summary metrics file}
PU_metrics <- read_sf(system.file("extdata", "targets.csv", package = "optimTFE"))
# get rid of any unneccessary columns
# metrics were calculated already from existing auxiliary datasets

# Merge to get all summary data and solution info, and spatial data. This could get
# big,so you may want to subset based on footprint unit count, etc.

# Keep unit_id and solution as columns 1 and 2, and columns for getting summary stats 
# (pu area, conservation lands, etc.)

sample_sols100_metrics <- left_join(PU_metrics, sample_sols100, by = c("PU_num" = "unit_id"))
# renaming so we can pull spatial data later if desired
sample_sols100_metrics_df <- as.data.frame(sample_sols100_metrics) |>
  relocate(solution, .after = PU_num) |>
  select(-c(select_order, geom))
# confirm no geometry or geom column

# Make a new df, summarized by solution number. Get solution-level metrics:
# example of creating summary metrics:
all_sols_metrics <- sample_sols100_metrics_df |>
  filter(!is.na(solution)) |> #some units do not occur in any solution
  group_by(solution) |>
  summarise(
    total_area_kmsq = sum(area_km2, na.rm = TRUE), # total footprint area
    good_hab_mean_pct = (sum(good_hab_kmsq, na.rm = TRUE) / sum(area_km2, na.rm 
          = TRUE)) * 100, #mean percent of footprint with 'good' habitat
    area_unfenced_mean_pct = (sum(not_fenced_kmsq, na.rm = TRUE) / sum(area_km2,
          na.rm = TRUE)) * 100, #mean percent of footprint unfenced
    area_fenced_mean_pct = (sum(fenced_kmsq, na.rm = TRUE) / sum(area_km2,
          na.rm = TRUE)) * 100,
    cons_lands_mean_pct = (sum(reserved_kmsq, na.rm = TRUE) / sum(area_km2, na.rm 
          = TRUE)) * 100,
    FB_richness_max_mean = mean(max_FB_richness, na.rm = TRUE),
    accessibility_max_mean = mean(max_accessibility, na.rm = TRUE)
  )
# Export as .csv for creating the optimTFE.explorer tool later. This is the column
# structure needed as input! 

write.csv(all_sols_metrics, file.path(path, "sample_sols100_metrics.csv" ))

```
## Open the explorer

```{r Open Explorer, warning=FALSE, message=FALSE}
explore_solutions()
```
